---
title: "Exercise 1 - Gonzalo Cardenal"
format: html
editor: visual
---

> ***TASK:*** Using quarto and R, create an executable HTML document with R code that samples 100 values from a log-normal distribution (see the rlnorm function; you can use mu=5, sigma=0.5); using the ggplot2 package, create a histogram of the sampled data on both the linear and log scales; create also a violin plot; within your quarto document, write 1-2 sentences to describe your steps and organize the document with subsection headings; report the mean and standard deviation of your sampled values in line in the text. Always show your code (i.e., do not hide when the HTML is built).

### Sampling values and basic statistics

We sample the values with the function *rlnorm* and report the mean and standard deviation.

```{r}
set.seed(12345)
samples <- rlnorm(100, meanlog = 5, sdlog = 0.5)
df <- data.frame(values = samples)
```

The mean of the sampled values is `r mean(samples)` and the standard deviation `r sd(samples)`.

### Plotting

We first load the library ggplot2

```{r}
library(ggplot2)
```

#### Histograms

Create the histogram in linear scale:

```{r}
linear_histogram <- ggplot(df, aes(x = samples)) + 
  geom_histogram(binwidth = 1, fill= "green", color="black") + 
  labs(title = "Histogram (Linear Scale)", x = "Values", y = "Frequency")
linear_histogram
```

and then we create the histogram in logaritmic scale:

```{r}
log_histogram <- ggplot(df, aes(x = log(samples))) +
  geom_histogram(binwidth = 0.7, fill = "blue", color="black")+
  scale_x_continuous(labels = scales::scientific, breaks = scales::trans_breaks("log10", function(x) 10^x)) +
  labs(title = "Histogram (Log Scale)", x = "Log Values", y = "Frequency")
log_histogram
```

finally the violin plot:

```{r}
violin_plot <- ggplot(df, aes(x="", y = samples)) +
  geom_violin(fill = "red", color = "black", alpha = 0.7) +
  labs(title = "Violin Plot", x = "", y = "Values")
violin_plot
```

and display the three plots together:

```{r}
library(gridExtra)
grid.arrange(linear_histogram, log_histogram, violin_plot, ncol = 2)
```
